---
title: "raptor survival using beta-binomial model"
---

```{r setup, include=FALSE}
library(rstan)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

```

Notes:  
* use subset of data from Newton et al. 2016 paper (adults only)  
* fixed effects: mass, type of estimate (direct/indirect survival based on methods used)  
* allowed intercept to vary per species, family and study  

I. Data Preparation  
A. Processed Raw Data  

```{r include=F}
surv=read.csv("https://raw.githubusercontent.com/patdumandan/RaptorSurvival/master/surv1.csv")

surv$mass=(surv$Average.mass..kg.-mean(surv$Average.mass..kg.))/(2*sd(surv$Average.mass..kg.))
surv$estimate<-ifelse(surv$death_type=="direct", 0, 1) #apparent (1) /true (0) survival estimate
surv$spcode=as.integer(surv$Species)
surv$famcode=as.integer(surv$family)
surv$stcode=as.integer(surv$Reference)

surv<-surv[which(surv$EnglishName!="Andean Condor"),] # remove Andean Condor because it is the only representative of its family, will screw up the LOO CV 
```

```{r echo=F}
str(surv)
```  

II. Data Analyses  
A. Beta-binomial model without specifications for LOO-CV post-processing  
**Notes**  
* This model was the one we agreed on that sort of works. We opted not to include diet and foraging strategies as covariates because Chris' original goal for this project was to predict survival based on mass, which is why we're doing the leave-one-out cross validation.  
* I'm not sure if I did the non-centered parameterization of the random intercepts correctly. 

```{r echo=F, message= F, warning=F}
dat_list=list( N=length(surv$Reference),
               y=surv$estimated.survived,    
               n=surv$sample.size,     
               mass=surv$mass,
               species=surv$spcode,
               family=surv$famcode,
               study=surv$stcode,
               Nsp=length(unique(surv$spcode)),
               Nst=length(unique(surv$stcode)),
               Nfam=length(unique(surv$famcode)),
               death_type=surv$estimate)
surv_mod=stan(model_code="

 data{

  int<lower=0> N; // no.of obs
  int <lower=0> y[N];       // survivors
  int <lower=0>  n[N];       // total 
  vector [N] mass;// ave.mass in kg
  int species[N]; //ID of each species
  int family [N]; //ID of family
  int study [N]; //ID of study
  int Nsp; //no.of species
  int Nst; //no.of studies
  int Nfam;// no. of families
  vector[N] death_type;// direct/indirect

 }
                
 parameters {

  real alpha;// global intercept
  real mass_eff; //slope for mass
  real est_eff; //slope for estimate type

  real<lower=0> sigma; //for NCP
  real <lower=0> sigma_sp[Nsp];//errors for random effects
  real<lower=0> sigma_st[Nst];//errors for random effects
  real<lower=0> sigma_fam[Nfam];//errors for random effects
 
  real <lower=0> phi;// variance of the likelihood
  real <lower=0, upper=1> pred_surv[N] ;//survival per observation
              }
   
     
   transformed parameters{
   
  vector <lower=0, upper=1> [N] surv_mu; //mean estimated survival 
  vector <lower=0> [N] A;
  vector <lower=0> [N] B;
  vector [Nsp] alpha_sp; //random intercept per species
  vector [Nst] alpha_st;// random intercept per study
  vector [Nfam] alpha_fam;// random intercept per family
 
  for (j in 1:Nsp) {
  
  alpha_sp[j]= alpha+sigma*sigma_sp[j];
  }

   for (k in 1:Nst) {
  
  alpha_st[k]= alpha+sigma*sigma_st[k];
   }
 
   for (m in 1:Nfam) {
  
 alpha_fam[m]= alpha+sigma*sigma_fam[m];
  }
 
  
  //model:
  
  for (i in 1:N){
  
  surv_mu[i]= inv_logit(alpha_sp[species[i]]+alpha_st[study[i]]+alpha_fam[family[i]]+
              mass_eff*mass[i]+est_eff*death_type[i]);

  }
  
  A = surv_mu * phi;
  B = (1 - surv_mu)* phi;
  
  }

 model {
  //priors

  mass_eff~ normal (0.1,1);
  est_eff~ normal (0,1);
 
  sigma~normal(0,1);
  sigma_sp~ normal(0,1);
  sigma_st~ normal(0,1);
  sigma_fam~ normal(0,1);
  
  phi ~normal(7,1);// use info. from beta regression of all juv and adult
  
  //model likelihood:
  
  pred_surv ~ beta(A, B); // survival estimate, beta dist.
  y~binomial(n, pred_surv); //no.of survivors drawn from binomial dist; based on sample size and reported survival estimate
 
  }", data=dat_list, chains=4, iter=3000)
```  

Model Output  
```{r echo=F}
post=extract(surv_mod)$pred_surv
matplot(surv$Average.mass..kg.,t(post), type="l",col="grey", xlab="average mass (kg)", ylab="survival estimate")
points(surv$survival.est~surv$Average.mass..kg., col="black", pch=19)
print(surv_mod, pars=c("alpha", "mass_eff", "alpha_sp", "alpha_fam", "alpha_st"))
```  

B. Beta-binomial model with the LOO-CV specifications  
**Notes**  
* This was where we left off last time and the reason why we removed Andean Condor because it would screw up predictions (only 1 representative of its family)  
* I am not totally sure what it is but it currently is not working well for me (it says I am accessing element out of range for alpha_st) 

Here, we created subsets of the data for training and predicting  
* surv_sub=subset of data that includes species that are not unique in the orig dataset(n=62) 
* surv_out=subset of data that includes species that are unique in the orig dataset(n=2)  
Then, we created separate "training" and "testing"(X_test) datasets using the surv_sub subset of data  

*Question*: If we did something using dplyr, would that have been problematic Something like:

```{r}
library(dplyr)
surv=surv%>%mutate(id=row_number())

train_dat=surv%>%sample_frac(0.8) #training dataset
test_dat=anti_join(surv, train_dat, by="id")
```
```{r}
dat_list_train<-vector("list",length=length(unique(surv$Species)))

dat_list_test<-vector("list",length=length(unique(surv$Species)))

for(i in 1:length(unique(surv$Species))){

surv_sub<-subset(surv,surv$Species!=unique(surv$Species)[i])
surv_out<-subset(surv,surv$Species==unique(surv$Species)[i])


dat_list_train[[i]]=list(
#training data
N=length(surv_sub$Species), y=surv_sub$estimated.survived, n=surv_sub$sample.size,
             diet=surv_sub$diet, forage=surv_sub$forage,
             mass=surv_sub$mass, death_type=surv_sub$estimate,
             species=surv_sub$spcode,family=surv_sub$famcode, study=surv_sub$stcode,
             Nfam=length(unique(surv_sub$family)), Nst=length(unique(surv_sub$Reference)), Nsp=length(unique(surv_sub$Species)),
             
#test data           
N_test=length(surv_sub$Species), n_test=surv_sub$sample.size,
             diet_test=surv_sub$diet, forage_test=surv_sub$forage,
             mass_test=surv_sub$mass, death_type_test=surv_sub$estimate,
             species_test=surv_sub$spcode,family_test=surv_sub$famcode, 
             study_test=surv_sub$stcode,Nfam_test=length(unique(surv_sub$family)),
             Nst_test=length(unique(surv_sub$Reference)), Nsp_test=length(unique(surv_sub$Species))
           
                   )

}
```  

```{r message=F, warning=F}
surv_mod_loocv=stan(model_code="

 data{

  int<lower=0> N; // no.of obs
  int <lower=0> y[N];       // survivors
  int <lower=0>  n[N];       // total 
  vector [N] mass;// ave.mass in kg
  int species[N]; //ID of each species
  int family [N]; //ID of family
  int study [N]; //ID of study
  int Nsp; //no.of species
  int Nst; //no.of studies
  int Nfam;// no. of families
  vector[N] death_type;// direct/indirect

//test data

  int<lower=0> N_test; // no.of obs
  int <lower=0>  n_test[N];       // total 
  vector [N_test] mass_test;// ave.mass in kg

  int species_test[N_test]; //ID of each species
  int family_test [N_test]; //ID of family
  int study_test [N_test]; //ID of study
  int Nsp_test; //no.of species
  int Nst_test; //no.of studies
  int Nfam_test;// no. of families
  vector[N_test] death_type_test;// direct/indirect

 }
                
 parameters {

  real alpha;// global intercept
  real mass_eff; //slope mass
  real est_eff; //slope indirect effect

  real<lower=0> sigma;//for NCP
  real<lower=0> sigma_sp[Nsp];//errors for random effects
  real<lower=0> sigma_st[Nst];//errors for random effects
  real<lower=0> sigma_fam[Nfam];//errors for random effects
  real <lower=0> phi;// variance for likelihood
  real <lower=0, upper=1> pred_surv[N] ;//survival per observation
  real <lower=0, upper=1> pred_surv_test[N_test];
              }
   
     
   transformed parameters{
  vector <lower=0, upper=1> [N] surv_mu; //mean estimated survival 
  vector <lower=0> [N] A;
  vector <lower=0> [N] B;
  vector [Nsp] alpha_sp; //random intercept per species
  vector [Nst] alpha_st;// random intercept per study
  vector [Nfam] alpha_fam;// random intercept per family

  
  for (j in 1:Nsp) {
  
  alpha_sp[j]= alpha+sigma*sigma_sp[j];
  }

   for (k in 1:Nst) {
  
  alpha_st[k]= alpha+sigma*sigma_st[k];
   }
 
   for (m in 1:Nfam) {
  
 alpha_fam[m]= alpha+sigma*sigma_fam[m];
 
  }
 
  
  //model:
  
  for (i in 1:N){
  
  surv_mu[i]= inv_logit(alpha_sp[species[i]]+alpha_st[study[i]]+alpha_fam[family[i]]+
  mass_eff*mass[i]+est_eff*death_type[i]);
  }
  
  A = surv_mu * phi;
  B = (1 - surv_mu)* phi;
  
  }

 model {
  //priors

  mass_eff~ normal (0.1,1);
  est_eff~ normal (0,1);
 
  sigma~ normal(0,1);
  sigma_sp~ normal(0,1);
  sigma_st~ normal(0,1);
  sigma_fam~ normal(0,1);
  
  phi ~normal(7,1);// use info. from beta regression of all juv and adult
  
  //model likelihood:
  
  pred_surv ~ beta(A, B); // survival estimate, beta dist.
  y~binomial(n, pred_surv); //no.of survivors drawn from binomial dist; based on sample size and reported survival estimate
 
  }

generated quantities {
  
  vector <lower=0, upper=1> [N_test] surv_mu_test;
  vector <lower=0> [N_test] A_test;
  vector <lower=0> [N_test] B_test;
  real pred_y_test [N_test];//predictions on survival
  int<lower=0> y_test[N_test]; //predictions on data level
 
  for (ii in 1:N_test){
  
  surv_mu_test[ii]= inv_logit(alpha_fam[family_test[ii]]+
  mass_eff*mass_test[ii]+est_eff*death_type_test[ii]);
  }
  
  A_test = surv_mu_test * phi;
  B_test = (1 - surv_mu_test)* phi;
  
    pred_y_test = beta_rng(A_test, B_test);
    y_test= binomial_rng(n_test, pred_y_test); 
}", 
data=dat_list_train[[i]], chains=4, iter=3000)
```  

Model Output  

```{r echo=F}
post1=extract(surv_mod_loocv)$pred_surv
matplot(surv$Average.mass..kg.,t(post1), type="l",col="grey", xlab="average mass (kg)", ylab="survival estimate")
points(surv$survival.est~surv$Average.mass..kg., col="black", pch=19)
print(surv_mod_loocv, pars=c("alpha", "mass_eff", "alpha_sp", "alpha_fam", "alpha_st"))
```  
